Fast and adaptive cointegration based model for forecasting financial time series

### 2017-12-11 slides/defense-slides.tex
First complete version for CCTVal

\documentclass{beamer}
%\usepackage[spanish,activeacute]{babel}
\usetheme{metropolis}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{array}
%\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{booktabs}
\usetikzlibrary{shapes,arrows}
%\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}

@@

\begin{frame}
\frametitle{Cointegration: formal definition}
\begin{block}{Cointegration definition}
Let {\color{blue}$\mathbf{y} = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$} be a set of $l$
time series of order I(1) which are said to be cointegrated if a vector
${\color{blue}\boldsymbol\beta=[\beta(1),\dots,\beta(l)]^\top} \in \mathbb{R}^l$  exists such that the
time series,

\begin{equation*}
 \mathbf{Z}_t:= \boldsymbol \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
 \text{I(0)}\, .
\end{equation*}
\end{block}
%In other words, a set of I(1) variables is said to be cointegrated if
%a linear combination of them exists which is I(0).

@@

{\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\mathbf{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}
}
\noindent where $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\top$. The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
\end{frame}

@@

@@

\begin{frame}
\frametitle{Data}
\begin{columns}
\begin{column}{0.7\textwidth}
\begin{itemize}
\item This data was collected from {\bf Dukascopy}, a free
database which gives access to the {\bf Swiss Foreign Exchange marketplace}.
\item Tests were carried out using four foreign exchange rates all related to
USD: {\bf EURUSD, GBPUSD, USDCHF} and {\bf USDJPY}. 
\item The tests were done using 10-seconds frequency from ask prices
from the 11th to the 15th of August 2014. 
%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
\end{itemize}
\end{column}
\begin{column}{0.3\textwidth}
 \begin{figure}[!h]
    \includegraphics[width=\textwidth]{img/dukascopy}
 \end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{PC vs MSE}
\begin{figure}[ht!]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.7\textwidth]{img/51_Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. }
  \label{fig:cointvsmse}
\end{figure}
\end{frame}

@@

\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
%\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
\item VECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
%\item Increasing the number of parameters will always lead to
%better performance measures.
\item The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds. 
\end{itemize}
\end{frame}

%\begin{frame}
%\frametitle{Proposal: Online VECM algorithm (OVECM)}
%\begin{itemize}
%\item Considers only a sliding window of the $L$ most recent data. 
%\item Since cointegration vectors represent long-run relationships they vary
%little in time. OVECM obtains these vectors only when they need to be updated. 
%\item This updating is done when MAPEs of the last $n$ inputs is above a certain
%error given by a threshold 
%\item OVECM also implements matrix optimisations in order to reduce execution time,
%such as updating matrices with new data, removing old data and introducing new
%cointegration vectors.
%\end{itemize}
%\end{frame}
%
%\begin{frame}
%\frametitle{Methodology}
%OVECM were compared against VECM and ARIMA. VECM and ARIMA are called with a
%sliding window of the most recent data, whereby the models are updated at every
%time step. Therefore, we named them SLVECM and SLARIMA.
%\end{frame}
%
%\subsection{Parameters}
%\begin{frame}
%\frametitle{Parameters optimization}
%VECM order and ARIMA parameters were selected using AIC.
%\begin{table}[ht]
%\label{tab:params}
%\begin{center}
%\begin{tabular}{|c|c|c|}
%\hline
%Windows size $L$ & VECM & ARIMA\\
% $L$ & order $(p)$ & order $(p,d,q)$ \\
%\hline
%100 & 2 & p=2,d=1,q=1\\
%400 & 5 & p=1,d=1,q=1\\
%700 & 3 &p=2,d=1,q=1\\
%1000 &3 & p=2,d=1,q=1\\
%\hline
%\end{tabular}
%\end{center}
%\end{table}
%\end{frame}
%
%\begin{frame}
%\frametitle{Execution times}
%We ran 400 iterations of OVECM and SLVECM . SLARIMA was based on the ARIMA procedure
%of the statsmodels library and therefore its execution time is excluded.
%%because its is not comparable with OVECM and SLVECM. SLARIMA was created based
%%on statsmodels library routine ARIMA.
%\begin{center}
%\adjustbox{max height=\dimexpr\textheight-5.5cm\relax,
%           max width=\textwidth}{
%\begin{tabular}{|l|c|c|c|c|c|}
%\hline
%& $L$ & order & e  & Time[s] \\
%\hline
%OVECM & 100 &p=2  & 0      & 2.492\\
%OVECM & 100 &p=2  & 0.0026  & 1.606\\
%SLVECM & 100 &p=2& -- & 2.100\\
%\hline
%OVECM & 400 & p=5  & 0      & 3.513\\
%OVECM & 400 &p=5  & 0.0041  & 2.569\\
%SLVECM & 400 & p=5 & -- & 3.222\\
%\hline
%OVECM & 700 &p=3  & 0      & 3.296\\
%OVECM & 700 &p=3  & 0.0032  & 2.856\\
%SLVECM & 700 &p=3 & -- & 3.581\\
%\hline
%OVECM & 1000 & p=3 & 0      & 4.387\\
%OVECM & 1000 & p=3  & 0.0022  & 2.408\\
%SLVECM & 1000 & p=3  & -- & 3.609\\
%\hline
%\end{tabular}
%}\end{center}
%\end{frame}
%
%\begin{frame}
%\frametitle{Performance accuracy}
%\tiny
%\begin{table}[H]
%\caption{Model measures}
%\label{tab:stats}
%\begin{center}
%\begin{tabular}{|l|l|c|c|c|c|c|c|c|}
%\hline
%\multicolumn{3}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
%\multicolumn{3}{|c|}{Out-of-sample} \\ 
%\hline
%\hline
%Method & $L$ & Parameters &  
%MAPE & MAE& RMSE& 
%MAPE & MAE& RMSE \\
%\hline
% OVECM  &   100  &  P=2 e=0.0026 &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
% OVECM  &   400  &  P=5 e=0.0041&  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
% OVECM  &   700  &  P=3 e=0.0032&  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
% OVECM  &   1000 &  P=3 e=0.0022&  
% \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
% \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
%\hline
% SLVECM  &   100 &  P=2 &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
% SLVECM  &   400 &  P=5 &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
% SLVECM  &   700 &  P=3 &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
% SLVECM  &   1000&  P=3 &  
% \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
% \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
%\hline
%SLARIMA & 100  &p=2 d=1 q=1 & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
%SLARIMA & 400  &p=1 d=1 q=1 & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
%SLARIMA & 700  &p=2 d=1 q=1 & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
%SLARIMA & 1000 &p=2 d=1 q=1 & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
%\hline
%\end{tabular}
%\end{center}
%\end{table}
%\end{frame}

\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item  OVECM considerably reduces execution times
without compromising solution accuracy.  
\item OVECM was compared with VECM and ARIMA
with the same sliding window sizes and OVECM outperformed both in terms of
execution time. 
\item Traditional VECM slightly outperformed our proposal but the
OVECM execution time is lower.
\item We could see that our algorithm took much less than a minute at every
step. This means that it could also be used with higher frequency data and would
still provide responses before new data arrives.  
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Summary of Contributions}
\begin{itemize}
\item Two different approaches based on cointegration were presented: an adaptive version of VECM and an online approach.
\item The concept of cointegration can now be used with high frequency time series.
\item VECM computational limitations can either be tackled using a distribution computations or considering an online approach.
\item An interesting connection between the percentage cointegration and forecasting performance.
\item Cointegration information can be used as an integration tool to detect arbitrage opportunities.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Author Contributions}
\begin{itemize}
\item 1
\item 2
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Future research}
\begin{itemize}
\item 1
\item 2
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{VECM(p) matrix form}
If we have $N$ data points, with $N>p$, we can estimate the model parameters $\mathbf{X}$ using the following matrix form:
\small
\begin{equation*}
\underbrace{
      \begin{bmatrix}
       \mathbf{\Delta y}_{p+1}  \\ 
       \mathbf{\Delta y}_{p+2}  \\ 
       \vdots                   \\ 
       \mathbf{\Delta y}_N      
      \end{bmatrix}}_{\mathbf{B} } =
\arraycolsep=1.4pt  
\underbrace{\begin{bmatrix}
   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
  \end{bmatrix}}_{\mathbf{X}^\top}
\underbrace{\begin{bmatrix}
 \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
 \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
 \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
 \vdots                  & \vdots                 & \ddots&  \vdots                   \\
 \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
 1                      & 1                       & \dots     & 1   
 \end{bmatrix}}_{\mathbf{A}^\top}
+
\underbrace{\begin{bmatrix}
              \mathbf{\epsilon}_{p+1} \\ 
              \mathbf{\epsilon}_{p+2} \\ 
              \vdots \\ 
              \mathbf{\epsilon}_N
             \end{bmatrix}}_{\mathbf{E}^\top}
\end{equation*}
\end{frame}

>>>>

\documentclass{beamer}
%\usepackage[spanish,activeacute]{babel}
\usetheme{metropolis}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{pmat}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{array}
%\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{booktabs}
\usetikzlibrary{shapes,arrows}
%\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}

@@

\begin{frame}
\frametitle{Cointegration: formal definition}
\begin{block}{Cointegration definition}
Let {\color{blue}$\mathbf{y} = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$} be a set of $l$
time series of order I(1) which are said to be cointegrated if a vector
${\color{blue}\boldsymbol\beta=[\beta(1),\dots,\beta(l)]^\intercal} \in \mathbb{R}^l$  exists such that the
time series,

\begin{equation*}
 \mathbf{Z}_t:= \boldsymbol \beta^\intercal \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
 \text{I(0)}\, .
\end{equation*}
\end{block}
%In other words, a set of I(1) variables is said to be cointegrated if
%a linear combination of them exists which is I(0).

@@

{\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\mathbf{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}
}
\noindent where $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$. The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
\end{frame}

@@

\begin{frame}
\frametitle{VECM(p) matrix form}
If we have $N$ data points, with $N>p$, the VECM matrix form is as follows:
\small
{\color{blue}
\begin{equation*}
\underbrace{
      \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\intercal \\
   \Delta\mathbf{y}_{p+2}^\intercal \\
   \vdots \\
   \Delta\mathbf{y}_N^\intercal
   \end{bmatrix}}_{\mathbf{Y} } =
\arraycolsep=1.4pt
\underbrace{
\begin{pmat}[{....|}]
   \mathbf{y}_p^\intercal \boldsymbol{\beta} & \Delta \mathbf{y}_p^\intercal & \Delta\mathbf{y}_{p-1}^\intercal & \dots 
                    & \Delta\mathbf{y}_2^\intercal & 1 \cr
   \mathbf{y}_{p+1}^\intercal  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\intercal & \Delta\mathbf{y}_p^\intercal & \dots
                       & \Delta\mathbf{y}_3^\intercal & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\intercal  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\intercal & \Delta\mathbf{y}_{N-2}^\intercal & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\intercal & 1 \cr
   \end{pmat}}_{\mathbf{X}}
\underbrace{
   \begin{bmatrix}
   \boldsymbol{\alpha}^\intercal \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\intercal
   \end{bmatrix}
}_{\mathbf{W}}
+
\underbrace{
\begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\intercal \\
   \boldsymbol{\epsilon}_{p+2}^\intercal \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\intercal \\
   \end{bmatrix}
}_{\mathbf{E}}
\end{equation*}}

We can estimate the model parameters $\mathbf{W}$ using OLS:
\begin{equation}
{\color{blue}
\mathbf{Y} = \mathbf{X}\mathbf{W}+\mathbf{E}
}
\end{equation}

\begin{frame}
\frametitle{Online version of VECM}
VECM can be solved using a sliding windows of data, the problem can be expressed as follows:

{\color{blue}
\[
\mathbf{X}(t) = 
\begin{bmatrix} 
\mathbf{x}_{t-L}^\intercal \\ \vdots \\ \mathbf{x}_t^\intercal
\end{bmatrix} \; , 
{\bf Y}(t) = \begin{bmatrix} \mathbf{y}^\intercal_{t-L} \\ \vdots \\ \mathbf{y}^\intercal_t \end{bmatrix} \; .
\]
}

The optimal solution using a OLS is then:

{\color{blue}
\begin{eqnarray}
{\bf W}(t)_*&=&({\bf X}(t)^\intercal{\bf X}(t))^{-1}{\bf X}(t)^\intercal{\bf Y}(t) \\
&=&\Big( \displaystyle \sum_{i=0}^L \mathbf{x}_{t-i}
\mathbf{x}_{t-i}^\intercal \Big)^{-1} \displaystyle \sum_{i=0}^L \mathbf{x}_{t-i} \mathbf{y}^\intercal_{t-i}
\end{eqnarray}
}
\end{frame}

\begin{frame}
\begin{algorithm}[H]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\{\mathbf{x}_1,\dots,\mathbf{x}_N \}$: $N$ input vectors \\
$\{\mathbf{y}_1,\dots,\mathbf{y}_N \}$: $N$ targets \\
$L$: sliding window size ($L<N$) \\
\ENSURE  $\,$ \\
$\{f(\mathbf{x}_{L+1}),\dots,f(\mathbf{x}_N) \}$: model predictions \\
\STATE Initialize $\mathbf{A}=\displaystyle \sum_{t=1}^L \mathbf{x}_t
\mathbf{x}_t^\intercal$
and $\mathbf{b}=\displaystyle \sum_{t=1}^L \mathbf{x}_t \mathbf{y}^\intercal_{t}$
\FOR { $t = L+1$ to $N$ }
        \STATE read new $\mathbf{x}_t$
        \STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal-
\mathbf{x}_{t-L-1} \mathbf{x}_{t-L-1}^\intercal  $
        \STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
\mathbf{A}^{-1}\mathbf{x}_t$
        \STATE Read new $y_t$
        \STATE $\mathbf{b} = \mathbf{b} + y_t \mathbf{x}_t$
\ENDFOR

\end{algorithmic}
\caption{Online Ordinary Least Squares}
\label{alg:SLAAR}
\end{algorithm}
\end{frame}

@@

\begin{frame}
\frametitle{Data}
\begin{columns}
\begin{column}{0.7\textwidth}
\begin{itemize}
\item This data was collected from {\bf Dukascopy}, a free
database which gives access to the {\bf Swiss Foreign Exchange marketplace}.
\item Tests were carried out using four foreign exchange rates all related to
USD: {\bf EURUSD, GBPUSD, USDCHF} and {\bf USDJPY}. 
\item The tests were done using {\bf 10-seconds frequency} from ask prices
from the 11th to the 15th of August 2014. 
%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
\end{itemize}
\end{column}
\begin{column}{0.3\textwidth}
 \begin{figure}[!h]
    \includegraphics[width=\textwidth]{img/dukascopy}
 \end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{PC vs MSE}
\begin{figure}[ht!]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.8\textwidth]{img/51_Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. }
  \label{fig:cointvsmse}
\end{figure}
\end{frame}

@@

\begin{frame}
\frametitle{Conclusions I}
\begin{itemize}
\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
%\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
\item AVECM improves performance measures by finding parameters of $L$ and $p$ maximising the percentage of cointegration.
\item AVECM performance is superior to ARIMA and the naive random walk model in terms of MSE and $U$-statistic. 
%This result is very important since it is still difficult to outperform ARIMA and the random walk model for standard econometric forecasting models  despite their simplicity.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Conclusions II}
\begin{itemize}
\item The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds.
\item We showed that VECM computational limitations can be tackled using a an online approach in a distributed environment.
\item Cointegration information can be used as an integration tool to detect arbitrage opportunities.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Future research}
\begin{itemize}
\item To explore the relationship between cointegration and performance for different assets (stocks, other forex rates).
\item To include more explicative variables such as bid-ask spread and change in volume.
\item The online approach for other econometrics models it will be also worthy of  study.% Many of them could be adapted and used with higher frequency data.
\item  In this thesis, the online version of OLS was studied, but the online version of
ridge regression could also be applied to obtain a solution with better
generalisation capabilities.
\end{itemize}
\end{frame}

@@
